---
title: "Data Processing"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
```{r, include = FALSE}
knitr::opts_chunk$set(
echo = TRUE,
message = FALSE,
warning = FALSE
)

library(tidyverse)
library(janitor)
```

# Source and study population

This project uses a de-identified, publicly available breast cancer dataset obtained from Kaggle (https://www.kaggle.com/datasets/amandam1/breastcancerdataset/code
). The dataset includes 334 patients with primary breast cancer who underwent surgical treatment. For each patient, the file contains demographic information (age, gender), tumour characteristics (tumour stage, histology, ER/PR/HER2 receptor status), several protein biomarkers (Protein1–Protein4), treatment information (type of surgery), and follow-up data (date of surgery, date of last visit, and vital status).

Before analysis, observations with missing patient_id were removed, as these corresponded to empty rows at the bottom of the file. For the survival analysis, we restricted to patients with non-missing dates of surgery and last follow-up, non-missing vital status, and non-negative follow-up time, resulting in 317 patients included in the Kaplan–Meier and log-rank analyses.

```{r, include = FALSE}
brca_df_raw = read_csv("BRCA.csv") |>
janitor::clean_names() |>
filter(!is.na(patient_id)) # remove empty rows at the bottom
```

# Variables

After cleaning the key columns are:

patient_id: Unique identifier for each patient.

age: Age at time of surgery (years), treated as a continuous covariate.

gender: Patient sex (factor).

protein1–protein4: Continuous protein biomarker measurements.

tumour_stage: Clinical tumour stage at diagnosis (I, II, III), treated as an ordered factor.

histology: Histological subtype of the breast tumour (categorical).

er_status: Oestrogen receptor (ER) status (e.g., Positive/Negative).

pr_status: Progesterone receptor (PR) status.

her2_status: HER2 receptor status.

surgery_type: Type of surgery performed (e.g., lumpectomy, mastectomy).

date_of_surgery: Calendar date of the primary breast cancer surgery.

date_of_last_visit: Date of last recorded follow-up visit.

patient_status: Vital status at last follow-up (Alive or Dead).

The primary outcome for this project is patient_status, indicating whether the patient was alive or dead at last follow-up. For survival analysis, we use date_of_surgery and date_of_last_visit to define follow-up time (in days) and combine this with patient_status to create a time-to-event and censoring indicator.

```{r, include = FALSE}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```

### Data import and cleaning

```{r, include=FALSE}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom
```

#### Variables

After `clean_names()`, your key columns are:

- `patient_id`: Unique identifier for each patient.
- `age`: Age at time of surgery (years), treated as a continuous covariate.
- `gender`: Patient sex (factor).
- `protein1`–`protein4`: Continuous protein biomarker measurements.
- `tumour_stage`: Clinical tumour stage at diagnosis (I, II, III), treated as an ordered factor.
- `histology`: Histological subtype of the breast tumour (categorical).
- `er_status`: Oestrogen receptor (ER) status (e.g., Positive/Negative).
- `pr_status`: Progesterone receptor (PR) status.
- `her2_status`: HER2 receptor status.
- `surgery_type`: Type of surgery performed (e.g., lumpectomy, mastectomy).

Time variables 

- Clarify outcome: patient status (Alive/Dead).
- `date_of_surgery`: Calendar date of the primary breast cancer surgery.
- `date_of_last_visit`: Date of last recorded follow-up visit.
- `patient_status`: Vital status at last follow-up (`Alive` or `Dead`).

The primary outcome for this project is **patient_status**, indicating whether the patient was alive or dead at last follow-up. For survival analysis, we use `date_of_surgery` and `date_of_last_visit` to define follow-up time (in days) and combine this with `patient_status` to create a time-to-event and censoring indicator.

#### Clean data types

```{r, include=FALSE}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```
