---
title: "Multiple Logistic Regression"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(broom)
library(ggplot2)
library(survminer)
library(pROC)        
library(survival)
library(dplyr)
library(modelr)
library(corrplot)
library(car)
```

```{r, include = FALSE}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom
```

```{r, include = FALSE}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```

# 4) Multiple logistic regression: Predicting patient status

### Model specification
- Outcome: patient_status (Alive vs Dead).
- Predictors: age, tumour_stage, surgery_type, selected proteins.
- Brief rationale for selected predictors (based on EDA).

### Model fitting
```{r}
brca =
  brca %>% 
  mutate(
    target = ifelse(patient_status == "Dead", 1, 0)
  )

brca_mlr =
  brca %>%  
  glm(target ~ age + tumour_stage + surgery_type + protein1 + protein2 +
        protein3 + protein4, data = .,
      family = binomial(link = "logit")
  )
```
We fit a multiple logistic regression to predict mortality using clinical predictors such as `age`, `tumour_stage`, `surgery_type`, and the protein biomarkers (`protein1`-`protein4`). These predictors were selected based on exploratory analyses showing differences in survival outcomes across different groups. 

### Model results
```{r}
brca_mlr %>% 
  broom::tidy(exponentiate = TRUE, conf.int = TRUE)
```
Model coefficient estimates were expressed as odds ratios. At the 0.05 level, none of the predictors were statistically significant, though the proteins generally expressed a weak positive association with mortality. Tumor stage showed a slight increase in odds of death, but the confidence intervals were wide. Overall, the clinical and protein predictors demonstrated a weak association with patient status.

### Model performance (ROC, AUC)
```{r}
roc_brca =
  brca %>% 
  mutate(pred_prob = predict(brca_mlr, newdata = ., type = "response")) %>% 
  roc(target, pred_prob) 

roc_brca %>% 
  ggroc() +
  labs(
    title = "ROC Curve for MLR Model",
    x = "False Positive Rate",
    y = "True Negative Rate"
  ) +
  theme_minimal()

auc(roc_brca)
```
The ROC curve suggests that the model was poor at distinguishing between patients who survived and passed away, with an AUC of `r round(auc(roc_brca), 3)`.

### Cross-validation
```{r}
set.seed(123)

cv_df = 
  crossv_mc(brca, 5) %>% 
  mutate(
    train = map(train, as_tibble),
    test  = map(test,  as_tibble)
  ) %>% 
  mutate(
    mlr_mod = map(
      .x = train,
      ~ glm(
          target ~ age + tumour_stage + surgery_type +
            protein1 + protein2 + protein3 + protein4,
          data   = .x,
          family = binomial(link = "logit")
        )
    ),
    pred_prob = map2(
      .x = mlr_mod,
      .y = test,
      ~ predict(.x, newdata = .y, type = "response")
    ),
    auc = map2_dbl(
      .x = pred_prob,
      .y = test,
      ~ pROC::auc(pROC::roc(.y$target, .x))
    )
  )

cv_summary =
  cv_df %>% 
  summarize(
    mean_auc = mean(auc),
    sd_auc   = sd(auc)
  )
```
To assess generalizability, we performed a 5-fold cross-validation. The average cross-validated AUC was `r round(cv_summary$mean_auc, 3)` with a standard deviation of `r round(cv_summary$sd_auc, 3)`. These values suggest that this model was minimally better than randomly predicting patient status.
