---
title: "Predicting Breast Cancer Outcomes from Clinical Features and Protein Biomarkers"
author: "Veerapetch Petchger, Clement Li, Heather Lu, Xiucheng (Eric) Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(broom)
library(pROC)        # ROC / AUC
library(survival)    # Surv, survfit
```

# 1) Introduction / Motivation (CLEMENT)

- Brief breast cancer background and burden.
- Why survival prediction and biomarkers matter.
- Your specific goals:
  - Describe clinical and biomarker patterns.
  - Build a logistic regression model for patient status.
  - Compare biomarker levels across groups.
  - Perform basic survival analysis.
  
# 2) Data

### Source and study population
- Kaggle dataset (334 breast cancer patients undergoing surgery).
- Link to source.
- Key inclusion/exclusion notes (if you drop rows).

### Data import and cleaning

```{r}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom
```

#### Variables

After `clean_names()`, your key columns are:

- `patient_id`
- `age`
- `gender`
- `protein1`–`protein4`
- `tumour_stage`
- `histology`
- `er_status`, `pr_status`, `her2_status`
- `surgery_type`
- `date_of_surgery`
- `date_of_last_visit`
- `patient_status`

- List key variables with short descriptions.

- Clarify outcome: patient status (Alive/Dead).

- Clarify time variables: Date_of_Surgery, Date_of_Last_Visit.

#### Clean data types

```{r}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```


## 3) Exploratory Data Analysis (EDA)

### Overall summaries
- Age: mean, SD, range.
- Tumour stage and histology: counts/percent.
- ER/PR/HER2 status: counts/percent.
- Surgery types: counts/percent.
- Protein biomarkers: mean, SD, distribution shape.

### Distributions

#### Age and proteins
- Histograms / density plots.
- Boxplots of proteins.

#### Stratified summaries
- Protein levels by tumour stage.
- Protein levels by histology.
- Crosstabs of receptor status and tumour stage.

```{r}

```


# 4) Logistic regression: Predicting patient status

### Model specification
- Outcome: patient_status (Alive vs Dead).
- Predictors: age, tumour_stage, surgery_type, ER/PR/HER2, selected proteins.
- Brief rationale for selected predictors (based on EDA).

### Model fitting
```{r}

```

### Model results
```{r}

```

### Model performance (ROC, AUC)
```{r}

```

# 5) ANOVA: Protein biomarkers across clinical groups (CLEMENT)
```{r}

```

## Question
- Do protein levels differ by tumour stage?
- Do protein levels differ by histology?

## Methods
- One-way ANOVA per protein.
- Post-hoc pairwise comparisons (if global p < 0.05).

## Results
- Brief table of means by group.
- ANOVA p-values.

```{r}

```

# 6) Survival analysis

### Time-to-event definition
- Follow-up time = Date_of_Last_Visit – Date_of_Surgery.
- Event indicator = 1 if Dead, 0 if Alive/unknown (define your handling of missing).

### Kaplan–Meier curves
- Overall KM curve.
- Stratified curves (e.g., tumour stage, HER2 status).

### Comparison
- Visual comparison.
- Optional: log-rank test.

# 7) Discussion (CLEMENT)

## Summary of main findings
- Key predictors from logistic regression.
- Whether biomarkers differ by stage/histology.
- Survival differences between groups.

## Strengths and limitations
- Small sample size (n = 334).
- Observational dataset; limited clinical detail.
- Missing data in follow-up and status.

## Future directions
- More advanced models (e.g., regularized logistic regression, Cox PH).
- External validation on independent data.
