---
title: "Predicting Breast Cancer Outcomes from Clinical Features and Protein Biomarkers"
author: "Veerapetch Petchger, Clement Li, Heather Lu, Xiucheng (Eric) Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(broom)
library(pROC)        # ROC / AUC
library(survival)    # Surv, survfit
```

# 1) Introduction / Motivation (CLEMENT)

- Brief breast cancer background and burden.
- Why survival prediction and biomarkers matter.
- Your specific goals:
  - Describe clinical and biomarker patterns.
  - Build a logistic regression model for patient status.
  - Compare biomarker levels across groups.
  - Perform basic survival analysis.
  
# 2) Data

### Source and study population
- Kaggle dataset (334 breast cancer patients undergoing surgery).
- Link to source.
- Key inclusion/exclusion notes (if you drop rows).

### Data import and cleaning

```{r}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom
```

#### Variables

After `clean_names()`, your key columns are:

- `patient_id`
- `age`
- `gender`
- `protein1`–`protein4`
- `tumour_stage`
- `histology`
- `er_status`, `pr_status`, `her2_status`
- `surgery_type`
- `date_of_surgery`
- `date_of_last_visit`
- `patient_status`

- List key variables with short descriptions.

- Clarify outcome: patient status (Alive/Dead).

- Clarify time variables: Date_of_Surgery, Date_of_Last_Visit.

#### Clean data types

```{r}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```


## 3) Exploratory Data Analysis (EDA)


## 1. Summary Tables

### Patient Characteristics Summary

```{r}
brca %>% 
  select(age, gender, tumour_stage, histology, er_status, pr_status, her2_status, surgery_type, patient_status) |> 
  skimr::skim()
```
### Protein Biomarker Summary
```{r}
brca %>% 
  summarise(
    across(starts_with("protein"), 
           list(mean = mean, sd = sd, min = min, max = max), 
           na.rm = TRUE)
  )
```
## 2. Distribution Visualization

### Age Distribution
```{r}
brca %>% 
  ggplot(aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Age Distribution of Breast Cancer Patients",
       x = "Age", y = "Count")
```
### Protein distribution
```{r}
brca %>%  
  pivot_longer(starts_with("protein"), names_to = "protein", values_to = "value") |> 
  ggplot(aes(x=value)) +
  geom_density(fill="pink", alpha=0.4) +
  facet_wrap(~protein, scales="free") +
  labs(title = "Distribution of Protein Biomarkers")

```

```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
```

## 3. Classification Visulization

### Carcinoma Type Classification
```{r}
carcinoma_type_pie =
  brca %>% 
  count(histology) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) %>% 
  ggplot(aes(x = "", y = prop, fill = histology)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Carcinoma / Histology Classification",
       fill  = "Histology") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

carcinoma_type_pie
```
Infiltrating ductal carcinoma (IDC) is the most common breast cancer type. It begins in the milk ducts and grows into the surrounding breast tissue. As it progresses, IDC can spread through the lymphatic system or bloodstream to other parts of the body.

Mucinous carcinoma is a breast cancer subtype in which tumor cells grow within mucin, the main component of mucus. Mucin normally supports healthy cell function, but in this tumor type, the mucin produced by cancer cells becomes part of the mass. Although mucinous tumors can occur elsewhere in the body, they most often appear in the breast.

Infiltrating lobular carcinoma (ILC) starts in the lobules, the glands that produce milk. Because it is an invasive cancer, it extends beyond the lobule into nearby tissue, and over time can spread to other parts of the body.

### Tumor Stage Classification
```{r}
tumour_stage_pie =
  brca %>% 
  count(tumour_stage) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) |>
  ggplot(aes(x = "", y = prop, fill = tumour_stage)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Tumour Stage Classification",
       fill  = "Tumour Stage") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

tumour_stage_pie
```



### General Patient Status (alive or dead)
```{r}
general_status_pie =
  brca %>% 
  count(patient_status) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) %>% 
  ggplot(aes(x = "", y = prop, fill = patient_status)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Patient Status",
       fill  = "Status") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

general_status_pie
```
### Patient status (alive or dead) based on tumor type
```{r}
histology_status_plot =
  brca %>% 
  count(histology, patient_status) %>% 
  group_by(histology) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(x = histology, y = prop, fill = patient_status)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Patient Status by Tumor Type (Histology)",
    x = "Histology",
    y = "Proportion within Histology",
    fill = "Patient status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

histology_status_plot
```
Mucinous carcinoma causes slightly more deaths in comparison to the other classifications of carcinoma.

### Patient status (alive or dead) based on tumor stage
```{r}
tumour_stage_status_plot =
  brca %>% 
  count(tumour_stage, patient_status) %>% 
  group_by(tumour_stage) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(x = tumour_stage, y = prop, fill = patient_status)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Patient Status by Tumour Stage",
    x = "Tumour stage",
    y = "Proportion within Stage",
    fill = "Patient status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

tumour_stage_status_plot
```
As expected, deaths rate increases with tumor stage. 


## 4. Stratified Summaries

### Protein levels by tumor stage
```{r}

```

### Protein levels by tumor type
```{r}

```



### Overall summaries
- Age: mean, SD, range.
- Tumour stage and histology: counts/percent.
- ER/PR/HER2 status: counts/percent.
- Surgery types: counts/percent.
- Protein biomarkers: mean, SD, distribution shape.

### Distributions

#### Age and proteins
- Histograms / density plots.
- Boxplots of proteins.

#### Stratified summaries
- Protein levels by tumour stage.
- Protein levels by histology.
- Crosstabs of receptor status and tumour stage.

```{r}

```


# 4) Logistic regression: Predicting patient status

### Model specification
- Outcome: patient_status (Alive vs Dead).
- Predictors: age, tumour_stage, surgery_type, ER/PR/HER2, selected proteins.
- Brief rationale for selected predictors (based on EDA).

### Model fitting
```{r}

```

### Model results
```{r}

```

### Model performance (ROC, AUC)
```{r}

```

# 5) ANOVA: Protein biomarkers across clinical groups (CLEMENT)
```{r}

```

## Question
- Do protein levels differ by tumour stage?
- Do protein levels differ by histology?

## Methods
- One-way ANOVA per protein.
- Post-hoc pairwise comparisons (if global p < 0.05).

## Results
- Brief table of means by group.
- ANOVA p-values.

```{r}

```

# 6) Survival analysis

### Time-to-event definition
- Follow-up time = Date_of_Last_Visit – Date_of_Surgery.
- Event indicator = 1 if Dead, 0 if Alive/unknown (define your handling of missing).

### Kaplan–Meier curves
- Overall KM curve.
- Stratified curves (e.g., tumour stage, HER2 status).

### Comparison
- Visual comparison.
- Optional: log-rank test.

# 7) Discussion (CLEMENT)

## Summary of main findings
- Key predictors from logistic regression.
- Whether biomarkers differ by stage/histology.
- Survival differences between groups.

## Strengths and limitations
- Small sample size (n = 334).
- Observational dataset; limited clinical detail.
- Missing data in follow-up and status.

## Future directions
- More advanced models (e.g., regularized logistic regression, Cox PH).
- External validation on independent data.
