---
title: "Predicting Breast Cancer Outcomes from Clinical Features and Protein Biomarkers"
author: "Veerapetch Petchger, Clement Li, Heather Lu, Xiucheng (Eric) Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(broom)
library(pROC)        # ROC / AUC
library(survival)    # Surv, survfit
```

# 1) Introduction / Motivation

- Brief breast cancer background and burden.
- Why survival prediction and biomarkers matter.
- Your specific goals:
  - Describe clinical and biomarker patterns.
  - Build a logistic regression model for patient status.
  - Compare biomarker levels across groups.
  - Perform basic survival analysis.
  
# 2) Data

### Source and study population
- Kaggle dataset (334 breast cancer patients undergoing surgery).
- Link to source.
- Key inclusion/exclusion notes (if you drop rows).

### Variables
- List key variables with short descriptions.
- Clarify outcome: patient status (Alive/Dead).
- Clarify time variables: Date_of_Surgery, Date_of_Last_Visit.

### Data import and cleaning

```{r}

```

## 3) Exploratory Data Analysis (EDA)

### Overall summaries
- Age: mean, SD, range.
- Tumour stage and histology: counts/percent.
- ER/PR/HER2 status: counts/percent.
- Surgery types: counts/percent.
- Protein biomarkers: mean, SD, distribution shape.

### Distributions

#### Age and proteins
- Histograms / density plots.
- Boxplots of proteins.

#### Stratified summaries
- Protein levels by tumour stage.
- Protein levels by histology.
- Crosstabs of receptor status and tumour stage.

```{r}

```


# 4) Logistic regression: Predicting patient status

### Model specification
- Outcome: patient_status (Alive vs Dead).
- Predictors: age, tumour_stage, surgery_type, ER/PR/HER2, selected proteins.
- Brief rationale for selected predictors (based on EDA).

### Model fitting
```{r}

```

### Model results
```{r}

```

### Model performance (ROC, AUC)
```{r}

```

# 5) ANOVA: Protein biomarkers across clinical groups
```{r}

```

## Question
- Do protein levels differ by tumour stage?
- Do protein levels differ by histology?

## Methods
- One-way ANOVA per protein.
- Post-hoc pairwise comparisons (if global p < 0.05).

## Results
- Brief table of means by group.
- ANOVA p-values.

```{r}

```

# 6) Survival analysis

### Time-to-event definition
- Follow-up time = Date_of_Last_Visit – Date_of_Surgery.
- Event indicator = 1 if Dead, 0 if Alive/unknown (define your handling of missing).

### Kaplan–Meier curves
- Overall KM curve.
- Stratified curves (e.g., tumour stage, HER2 status).

### Comparison
- Visual comparison.
- Optional: log-rank test.

# 7) Discussion

## Summary of main findings
- Key predictors from logistic regression.
- Whether biomarkers differ by stage/histology.
- Survival differences between groups.

## Strengths and limitations
- Small sample size (n = 334).
- Observational dataset; limited clinical detail.
- Missing data in follow-up and status.

## Future directions
- More advanced models (e.g., regularized logistic regression, Cox PH).
- External validation on independent data.
