---
title: "Predicting Breast Cancer Outcomes from Clinical Features and Protein Biomarkers"
author: "Veerapetch Petchger, Clement Li, Heather Lu, Xiucheng (Eric) Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

library(tidyverse)
library(janitor)
library(broom)
library(ggplot2)
library(survminer)
library(pROC)        
library(survival)
library(dplyr)
```

# 1) Introduction / Motivation (CLEMENT)

Breast cancer is a major global health burden and a leading cause of cancer-related death among women. While clinical characteristics such as tumor stage and histology inform prognosis, molecular biomarkers may offer additional insight into disease progression and survival. In this project, we describe patterns of clinical features and protein biomarkers in the BRCA dataset, assess whether biomarker levels differ across patient groups, build a logistic regression model to predict survival status, and conduct basic survival analysis to explore factors associated with follow-up outcomes.

# 2) Data

### Source and study population
- Kaggle dataset (334 breast cancer patients undergoing surgery).
- Link to source.
- Key inclusion/exclusion notes (if you drop rows).

### Data import and cleaning

```{r}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom
```

#### Variables

After `clean_names()`, your key columns are:

- `patient_id`
- `age`
- `gender`
- `protein1`–`protein4`
- `tumour_stage`
- `histology`
- `er_status`, `pr_status`, `her2_status`
- `surgery_type`
- `date_of_surgery`
- `date_of_last_visit`
- `patient_status`

- List key variables with short descriptions.

- Clarify outcome: patient status (Alive/Dead).

- Clarify time variables: Date_of_Surgery, Date_of_Last_Visit.

#### Clean data types

```{r}
brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```


## 3) Exploratory Data Analysis (EDA)


## 1. Summary Tables

### Patient Characteristics Summary

```{r}
brca %>% 
  select(age, gender, tumour_stage, histology, er_status, pr_status, her2_status, surgery_type, patient_status) |> 
  skimr::skim()
```
### Protein Biomarker Summary
```{r}
brca %>% 
  summarise(
    across(starts_with("protein"), 
           list(mean = mean, sd = sd, min = min, max = max), 
           na.rm = TRUE)
  )
```
## 2. Distribution Visualization

### Age Distribution
```{r}
brca %>% 
  ggplot(aes(x = age)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(title = "Age Distribution of Breast Cancer Patients",
       x = "Age", y = "Count")
```
### Protein distribution
```{r}
brca %>%  
  pivot_longer(starts_with("protein"), names_to = "protein", values_to = "value") |> 
  ggplot(aes(x=value)) +
  geom_density(fill="pink", alpha=0.4) +
  facet_wrap(~protein, scales="free") +
  labs(title = "Distribution of Protein Biomarkers")

```

## 3. Classification Visulization

### Carcinoma Type Classification
```{r}
carcinoma_type_pie =
  brca %>% 
  count(histology) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) %>% 
  ggplot(aes(x = "", y = prop, fill = histology)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Carcinoma / Histology Classification",
       fill  = "Histology") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

carcinoma_type_pie
```
Infiltrating ductal carcinoma (IDC) is the most common breast cancer type. It begins in the milk ducts and grows into the surrounding breast tissue. As it progresses, IDC can spread through the lymphatic system or bloodstream to other parts of the body.

Mucinous carcinoma is a breast cancer subtype in which tumor cells grow within mucin, the main component of mucus. Mucin normally supports healthy cell function, but in this tumor type, the mucin produced by cancer cells becomes part of the mass. Although mucinous tumors can occur elsewhere in the body, they most often appear in the breast.

Infiltrating lobular carcinoma (ILC) starts in the lobules, the glands that produce milk. Because it is an invasive cancer, it extends beyond the lobule into nearby tissue, and over time can spread to other parts of the body.

### Tumor Stage Classification
```{r}
tumour_stage_pie =
  brca %>% 
  count(tumour_stage) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) |>
  ggplot(aes(x = "", y = prop, fill = tumour_stage)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Tumour Stage Classification",
       fill  = "Tumour Stage") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

tumour_stage_pie
```



### General Patient Status (alive or dead)
```{r}
general_status_pie =
  brca %>% 
  count(patient_status) %>% 
  mutate(
    prop  = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  ) %>% 
  ggplot(aes(x = "", y = prop, fill = patient_status)) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(title = "Patient Status",
       fill  = "Status") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5))

general_status_pie
```
### Patient status (alive or dead) based on tumor type
```{r}
histology_status_plot =
  brca %>% 
  count(histology, patient_status) %>% 
  group_by(histology) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(x = histology, y = prop, fill = patient_status)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Patient Status by Tumor Type (Histology)",
    x = "Histology",
    y = "Proportion within Histology",
    fill = "Patient status"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )

histology_status_plot
```
Mucinous carcinoma causes slightly more deaths in comparison to the other classifications of carcinoma.

### Patient status (alive or dead) based on tumor stage
```{r}
tumour_stage_status_plot =
  brca %>% 
  count(tumour_stage, patient_status) %>% 
  group_by(tumour_stage) %>% 
  mutate(prop = n / sum(n)) %>% 
  ggplot(aes(x = tumour_stage, y = prop, fill = patient_status)) +
  geom_col(position = "fill", color = "white") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Patient Status by Tumour Stage",
    x = "Tumour stage",
    y = "Proportion within Stage",
    fill = "Patient status"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5)
  )

tumour_stage_status_plot
```
As expected, deaths rate increases with tumor stage. 


## 4. Stratified Summaries

### Protein levels by tumor stage
```{r}

```

### Protein levels by tumor type
```{r}

```



### Overall summaries
- Age: mean, SD, range.
- Tumour stage and histology: counts/percent.
- ER/PR/HER2 status: counts/percent.
- Surgery types: counts/percent.
- Protein biomarkers: mean, SD, distribution shape.

### Distributions

#### Age and proteins
- Histograms / density plots.
- Boxplots of proteins.

#### Stratified summaries
- Protein levels by tumour stage.
- Protein levels by histology.
- Crosstabs of receptor status and tumour stage.

```{r}

```


# 4) Logistic regression: Predicting patient status

### Model specification
- Outcome: patient_status (Alive vs Dead).
- Predictors: age, tumour_stage, surgery_type, ER/PR/HER2, selected proteins.
- Brief rationale for selected predictors (based on EDA).

### Model fitting
```{r}

```

### Model results
```{r}

```

### Model performance (ROC, AUC)
```{r}

```

# 5) ANOVA: Protein biomarkers across clinical groups (CLEMENT)
```{r}
# identify protein biomarker columns (protein1–protein4)
protein_vars <- names(brca)[grepl("^protein", names(brca))]
```

##### Do protein levels differ by tumour stage?
```{r}
# group means by tumour_stage
means_tumour_stage <- brca |>
  select(tumour_stage, all_of(protein_vars)) |>
  pivot_longer(
    cols = all_of(protein_vars),
    names_to = "biomarker",
    values_to = "value"
  ) |>
  group_by(tumour_stage, biomarker) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    .groups = "drop"
  )

# one-way ANOVA for each protein ~ tumour_stage
anova_tumour_stage <- protein_vars |>
  lapply(function(p) {
    fit <- aov(reformulate("tumour_stage", response = p), data = brca)
    tidy(fit) |>
      slice(1) |>
      transmute(
        biomarker = p,
        df = df,
        f_statistic = statistic,
        p_value = p.value
      )
  }) |>
  bind_rows() |>
  mutate(p_adj = p.adjust(p_value, method = "BH"))

means_tumour_stage
anova_tumour_stage
```

For all four proteins (protein1–protein4), the global ANOVA tests were not statistically significant (all BH-adjusted p-values > 0.77). Thus, there is no statistical evidence of differences in biomarker expression by tumour stage in this cohort, and post-hoc pairwise comparisons were not pursued. Although small numerical differences in mean protein levels were observed, these differences were not statistically significant. This suggests that, in this dataset, tumour stage alone does not explain variation in protein biomarker expression. This could be due to biological similarity across stages, high within-group variability, or limited power for detecting subtle effects.

##### Do protein levels differ by histology?
```{r}
# group means by histology
means_histology <- brca |>
  select(histology, all_of(protein_vars)) |>
  pivot_longer(
    cols = all_of(protein_vars),
    names_to = "biomarker",
    values_to = "value"
  ) |>
  group_by(histology, biomarker) |>
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    n = sum(!is.na(value)),
    .groups = "drop"
  )

# one-way ANOVA for each protein ~ histology
anova_histology <- protein_vars |>
  lapply(function(p) {
    fit <- aov(reformulate("histology", response = p), data = brca)
    tidy(fit) |>
      slice(1) |>
      transmute(
        biomarker = p,
        df = df,
        f_statistic = statistic,
        p_value = p.value
      )
  }) |>
  bind_rows() |>
  mutate(p_adj = p.adjust(p_value, method = "BH"))

means_histology
anova_histology
```

Across histologic subtypes, protein levels were visually and statistically similar, with substantial overlap between groups.
All BH-adjusted ANOVA p-values were > 0.36, indicating no detectable expression differences by histology.
Consequently, post-hoc pairwise comparisons were not performed.

Because no p-values are statistically significant, we check ANOVA assumptions 

##### Normality (per group per protein)
```{r}
for (p in protein_vars) {
print(
ggqqplot(brca, x = p, facet.by = "tumour_stage",
title = paste("QQ-Plot:", p, "by tumour stage"))
)
}
```

The QQ-plots for each protein by tumour stage were reasonably linear, suggesting that the normality assumption is acceptable for these ANOVA models.

##### Homogeneity of variances (Levene’s Test)
```{r}
library(car)

for (p in protein_vars) {
  print(p)
  print(leveneTest(brca[[p]] ~ brca$tumour_stage))
}
```

Levene’s tests indicated no significant variance differences across tumour stage groups for any protein (all p > 0.18).
Thus, the assumption of equal variances is not violated for any protein, and standard one-way ANOVA is appropriate and valid. Overall, protein levels do not significantly differ by tumour stage in this dataset — not because ANOVA assumptions failed, but because any true effects are likely small relative to within-group variability.

##### Boxplot — Proteins by Tumour Stage
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Identify proteins
protein_vars <- names(brca)[grepl("^protein", names(brca))]

# Pivot long
brca_long_stage <- brca |>
  select(tumour_stage, all_of(protein_vars)) |>
  filter(!is.na(tumour_stage)) |>
  pivot_longer(
    cols = all_of(protein_vars),
    names_to = "protein",
    values_to = "value"
  )

# Boxplots 
ggplot(brca_long_stage, aes(x = tumour_stage, y = value, fill = tumour_stage)) +
  geom_boxplot(outlier.alpha = 0.2, width = 0.6) +
  geom_jitter(alpha = 0.4, width = 0.15, size = 1) +
  facet_wrap(~ protein, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Protein Levels by Tumour Stage",
    x = "Tumour Stage",
    y = "Protein Expression"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

```

Protein1 expression showed very similar distributions across tumour stages I, II, and III. Median levels were close to zero in all three groups, and the interquartile ranges largely overlapped. Although a few outliers were present in each stage, the overall spread of values was comparable, suggesting no meaningful differences in Protein1 expression by tumor stage in this dataset. 
This interpretation perfectly matches our ANOVA result where p = 0.50 (not significant) and Levene’s test where p = 0.18 (equal variances). Similar patterns were observed for protein2–protein4, with substantial overlap in distributions across tumor stages.

##### Boxplot - Protein by Histology
```{r}
# Identify protein variables
protein_vars <- names(brca)[grepl("^protein", names(brca))]

# Pivot to long format
brca_long_hist <- brca |>
  select(histology, all_of(protein_vars)) |>
  filter(!is.na(histology)) |>
  pivot_longer(
    cols = all_of(protein_vars),
    names_to = "protein",
    values_to = "value"
  )

# Plot
ggplot(brca_long_hist, aes(x = histology, y = value, fill = histology)) +
  geom_boxplot(outlier.alpha = 0.2, width = 0.6) +
  geom_jitter(alpha = 0.4, width = 0.15, size = 1) +
  facet_wrap(~ protein, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Protein Levels by Histology",
    x = "Histologic Subtype",
    y = "Protein Expression"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold")
  )

```

Across histologic subtypes, protein levels were visually similar with substantial overlap between groups, indicating no detectable expression differences by histology.

This is consistent with the non-significant ANOVA results (all adjusted p-values > 0.36).

# 6) Survival analysis

### Time-to-event definition
- Follow-up time = Date_of_Last_Visit – Date_of_Surgery.
- Event indicator = 1 if Dead, 0 if Alive/unknown (define your handling of missing).

```{r}
brca_surv = brca |>
  filter(!is.na(event), !is.na(follow_up_time), follow_up_time >= 0)

# Create Survival object
surv_obj = Surv(time = brca_surv$follow_up_time,
                event = brca_surv$event)
```

### Kaplan–Meier curves
- Overall KM curve.
- Stratified curves (e.g., tumour stage, HER2 status).

#### Kaplan–Meier curves

An overall Kaplan–Meier curve was estimated to describe the distribution of survival time in the cohort. The plot displays the estimated survival probability over follow-up, with 95% confidence bands and a risk table indicating the number of patients at risk at selected time points.

```{r}
km_overall = survfit(surv_obj ~ 1, data = brca_surv)

ggsurvplot(
  km_overall,
  conf.int = TRUE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  surv.median.line = "hv",
  risk.table = TRUE
)

```

```{r}
km_stage = survfit(surv_obj ~ tumour_stage, data = brca_surv)

ggsurvplot(
  km_stage,
  data = brca_surv,
  conf.int = FALSE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  legend.title = "Tumour stage",
  risk.table = TRUE,
  pval = TRUE         # shows log-rank p-value on the plot
)

```

```{r}
km_her2 = survfit(surv_obj ~ her2_status, data = brca_surv)

ggsurvplot(
  km_her2,
  data = brca_surv,
  conf.int = FALSE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  legend.title = "HER2 status",
  risk.table = TRUE,
  pval = TRUE
)

```

Kaplan–Meier curves were also stratified by tumour stage (I, II, III) and by HER2 receptor status. These plots allow visual comparison of survival across subgroups. In both cases, a risk table was included to show how many individuals remained under observation at different time points.

### Comparison

```{r}
# Log-rank test: tumour stage
logrank_stage = survdiff(surv_obj ~ tumour_stage, data = brca_surv)
logrank_stage

# Log-rank test: HER2 status
logrank_her2 = survdiff(surv_obj ~ her2_status, data = brca_surv)
logrank_her2
```
### Results

##### Overall survival

The overall Kaplan–Meier curve shows a gradual decline in survival probability over the follow-up period. Most patients remain alive during the first part of follow-up, and the curve flattens out after roughly 4 years, indicating that a sizeable fraction of patients survive long-term. The median survival time is approximately 1500 days, meaning that half of the cohort is expected to be alive beyond this time point.The step-wise drops in the curve correspond to observed deaths, and the tick marks indicate censored observations (patients last seen alive at that time).

##### Survival by tumour stage

Kaplan–Meier curves stratified by tumour stage (I, II, III) suggest that patients with more advanced disease tend to have poorer survival (Figure Y). Stage I patients generally have the highest estimated survival probabilities throughout follow-up, Stage II patients are intermediate, and Stage III patients have the lowest survival, especially after about 1,000 days.

A log-rank test comparing the three stages gives a p-value of 0.10. This is larger than the conventional 0.05 threshold, so there is no strong statistical evidence of a difference in survival across tumour stages in this dataset. However, the pattern of the curves is consistent with the expected clinical trend of worse survival for higher stage disease.

##### Survival by HER2 status

Kaplan–Meier curves stratified by HER2 status (positive vs negative) are almost completely overlapping (Figure Z). The log-rank test yields a p-value of 0.97, indicating no evidence of a difference in survival between HER2-positive and HER2-negative patients in this sample. This result should be interpreted cautiously because the HER2-positive group is relatively small (29 patients vs 288 HER2-negative), so the comparison may have limited power to detect modest differences.

# 7) Discussion (CLEMENT)

## Summary of main findings
- Key predictors from logistic regression.
- Whether biomarkers differ by stage/histology.
- Survival differences between groups.

## Strengths and limitations
- Small sample size (n = 334).
- Observational dataset; limited clinical detail.
- Missing data in follow-up and status.

## Future directions
- More advanced models (e.g., regularized logistic regression, Cox PH).
- External validation on independent data.
