---
title: "Survival Analysis"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(janitor)
library(broom)
library(ggplot2)
library(survminer)
library(pROC)        
library(survival)
library(dplyr)
library(modelr)
library(corrplot)
library(car)
```

```{r, include=FALSE}
brca_df_raw = read_csv("BRCA.csv") |> 
  janitor::clean_names() |> 
  filter(!is.na(patient_id)) #remove empty rows at the bottom

brca = brca_df_raw |> 
  mutate( #categorical variables
    gender       = factor(gender),
    tumour_stage = factor(tumour_stage,
                          levels = c("I", "II", "III"),
                          ordered = TRUE),
    histology    = factor(histology),
    er_status    = factor(er_status),
    pr_status    = factor(pr_status),
    her2_status  = factor(her2_status),
    surgery_type = factor(surgery_type),
    patient_status = factor(patient_status,
                            levels = c("Alive", "Dead")),
    #date variables
    date_of_surgery     = as.Date(date_of_surgery, format = "%d-%b-%y"),
    date_of_last_visit  = as.Date(date_of_last_visit, format = "%d-%b-%y"),

    # survival
    follow_up_time = as.numeric(date_of_last_visit - date_of_surgery),
    event = case_when(
      patient_status == "Dead"  ~ 1L,
      patient_status == "Alive" ~ 0L,
      TRUE                      ~ NA_integer_
    )
  )
```

# 6) Survival analysis

### Time-to-event definition
- Follow-up time = Date_of_Last_Visit – Date_of_Surgery.
- Event indicator = 1 if Dead, 0 if Alive/unknown 

```{r}
brca_surv = brca |>
  filter(!is.na(event), !is.na(follow_up_time), follow_up_time >= 0)

# Create Survival object
surv_obj = Surv(time = brca_surv$follow_up_time,
                event = brca_surv$event)
```

### Kaplan–Meier curves
- Overall KM curve.
- Stratified curves (e.g., tumour stage, HER2 status).

#### Kaplan–Meier curves

An overall Kaplan–Meier curve was estimated to describe the distribution of survival time in the cohort. The plot displays the estimated survival probability over follow-up, with 95% confidence bands and a risk table indicating the number of patients at risk at selected time points.

```{r}
km_overall = survfit(surv_obj ~ 1, data = brca_surv)

ggsurvplot(
  km_overall,
  conf.int = TRUE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  surv.median.line = "hv",
  risk.table = TRUE
)

```

```{r}
km_stage = survfit(surv_obj ~ tumour_stage, data = brca_surv)

ggsurvplot(
  km_stage,
  data = brca_surv,
  conf.int = FALSE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  legend.title = "Tumour stage",
  risk.table = TRUE,
  pval = TRUE         # shows log-rank p-value on the plot
)

```

```{r}
km_her2 = survfit(surv_obj ~ her2_status, data = brca_surv)

ggsurvplot(
  km_her2,
  data = brca_surv,
  conf.int = FALSE,
  xlab = "Follow-up time (days)",
  ylab = "Survival probability",
  legend.title = "HER2 status",
  risk.table = TRUE,
  pval = TRUE
)

```

Kaplan–Meier curves were also stratified by tumour stage (I, II, III) and by HER2 receptor status. These plots allow visual comparison of survival across subgroups. In both cases, a risk table was included to show how many individuals remained under observation at different time points.

### Comparison

```{r}
# Log-rank test: tumour stage
logrank_stage = survdiff(surv_obj ~ tumour_stage, data = brca_surv)
logrank_stage

# Log-rank test: HER2 status
logrank_her2 = survdiff(surv_obj ~ her2_status, data = brca_surv)
logrank_her2
```
### Results

##### Overall survival

The overall Kaplan–Meier curve shows a gradual decline in survival probability over the follow-up period. Most patients remain alive during the first part of follow-up, and the curve flattens out after roughly 4 years, indicating that a sizeable fraction of patients survive long-term. The median survival time is approximately 1500 days, meaning that half of the cohort is expected to be alive beyond this time point.The step-wise drops in the curve correspond to observed deaths, and the tick marks indicate censored observations (patients last seen alive at that time).

##### Survival by tumour stage

Kaplan–Meier curves stratified by tumour stage (I, II, III) suggest that patients with more advanced disease tend to have poorer survival (Figure Y). Stage I patients generally have the highest estimated survival probabilities throughout follow-up, Stage II patients are intermediate, and Stage III patients have the lowest survival, especially after about 1,000 days.

A log-rank test comparing the three stages gives a p-value of 0.10. This is larger than the conventional 0.05 threshold, so there is no strong statistical evidence of a difference in survival across tumour stages in this dataset. However, the pattern of the curves is consistent with the expected clinical trend of worse survival for higher stage disease.

##### Survival by HER2 status

Kaplan–Meier curves stratified by HER2 status (positive vs negative) are almost completely overlapping (Figure Z). The log-rank test yields a p-value of 0.97, indicating no evidence of a difference in survival between HER2-positive and HER2-negative patients in this sample. This result should be interpreted cautiously because the HER2-positive group is relatively small (29 patients vs 288 HER2-negative), so the comparison may have limited power to detect modest differences.
